Carryover effects in free recall reveal how prior experiences influence memories of new experiences

### 2022-12-29 paper/main.tex
checking in first draft of intro

Over the past half-century, context-based models have enjoyed impressive
success at explaining many stereotyped behaviors observed during free recall
and other list-learning tasks~\citep{Este55a, RaaiShif80, GlenEtal83,
HowaKaha02a, SiroEtal05, KimbEtal07, PolyKaha08, PolyEtalTulv, SedeEtal08,
PolyEtal09, ShanHowa12}. These phenomena include the well-known recency and
primacy effects (superior recall of items from the end and, to a lesser extent,
from the beginning of the study list), as well as semantic and temporal
clustering effects~\citep{KahaEtal08}. The contiguity effect is an example of
temporal clustering, which is perhaps the dominant form of organization in free
recall. This effect can be seen in the tendency for people to successively
recall items that occupied neighboring positions in the study list. For
example, if a list contained the sub-sequence \textsc{``absence hollow pupil''}
and the participant recalls the word \textsc{``hollow''}, it is far more likely
that the next response will be either \textsc{``pupil''} or
\textsc{``absence''} than some other list item~\citep{Kaha96}. In addition,
there is a strong forward bias in the contiguity effect: subjects make forward
transitions (i.e., \textsc{``hollow''} followed by \textsc{``pupil''}) about
twice as often as they make backward transitions, despite an overall tendency
to begin recall at the end of the list. There are also striking effects of
semantic clustering~\citep{RomnEtal93, Bous53, BousEtal54, JenkRuss52,
MannKaha12}, whereby the recall of a given item is more likely to be followed
by recall of a similar or related item than a dissimilar or unrelated one.

In general, people organize memories for words along a wide variety of stimulus
dimensions. As captured by models like the \textit{Context Maintenance and
Retrieval Model}~\citep{PolyEtal09}, the stimulus features associated with each
word (e.g.\ the word's meaning, font size, font color, location on the screen,
size of the object the word represents, etc.) are incorporated into the
participant's mental context representation~\citep{SmitVela01, MannEtal15, Mann20}.
During a memory test, any of these features may serve as a memory cue, which in
turn leads the participant to recall in succession words that share stimulus
features.

@@

These clustering behaviors in ....

% feature-rich free recall, basic manipulation conditions, preview of findings

>>>>

Over the past half-century, context-based models have enjoyed impressive
success at explaining many stereotyped behaviors observed during free recall
and other list-learning tasks~\citep{Este55a, RaaiShif80, GlenEtal83,
HowaKaha02a, SiroEtal05, KimbEtal07, PolyKaha08, PolyEtalTulv, SedeEtal08,
PolyEtal09, ShanHowa12}. These phenomena include the well-known recency and
primacy effects (superior recall of items from the end and, to a lesser extent,
from the beginning of the study list), as well as semantic and temporal
clustering effects~\citep{KahaEtal08}. The contiguity effect is an example of
temporal clustering, which is perhaps the dominant form of organization in free
recall. This effect can be seen in the tendency for people to successively
recall items that occupied neighboring positions in the study list. For
example, if a list contained the sub-sequence \textsc{``absence hollow pupil''}
and the participant recalls the word \textsc{``hollow''}, it is far more likely
that the next response will be either \textsc{``pupil''} or
\textsc{``absence''} than some other list item~\citep{Kaha96}. In addition,
there is a strong forward bias in the contiguity effect: subjects make forward
transitions (i.e., \textsc{``hollow''} followed by \textsc{``pupil''}) about
twice as often as they make backward transitions, despite an overall tendency
to begin recall at the end of the list. There are also striking effects of
semantic clustering~\citep{RomnEtal93, Bous53, BousEtal54, JenkRuss52,
MannKaha12}, whereby the recall of a given item is more likely to be followed
by recall of a similar or related item than a dissimilar or unrelated one. In
general, people organize memories for words along a wide variety of stimulus
dimensions. As captured by models like the \textit{Context Maintenance and
Retrieval Model}~\citep{PolyEtal09}, the stimulus features associated with each
word (e.g.\ the word's meaning, font size, font color, location on the screen,
size of the object the word represents, etc.) are incorporated into the
participant's mental context representation~\citep{SmitVela01, MannEtal15,
Mann20, MannEtal11, MannEtal12}. During a memory test, any of these features
may serve as a memory cue, which in turn leads the participant to recall in
succession words that share stimulus features.

@@

A key mystery is whether the sorts of situation models and schemas that people
use to organize their memories of real-world experiences might map onto the
clustering effects that reflect how people organize their memories for word
lists. On one hand, situation models and clustering effects both reflect
statistical regularities in ongoing experience. Our memory systems exploit
these regularities when generating inferences about the unobserved past and
yet-to-be-experienced future~\citep{XuEtal22, SchaTurk15, RangRitc12,
BoweEtal79, MomeEtal17}. On the other hand, the rich structure of real-world
experiences and other naturalistic stimuli that enable people to form deep and meaningful
situation models and schemas have no obvious analog in simple word lists.  Often lists
in free recall studies are explicitly \textit{designed} to be devoid of exploitable
temporal structure, for example by sorting the words in a random order~\citep{Kaha12}.

% feature-rich free recall, basic manipulation conditions, preview of findings
We designed an experimental paradigm to explore how people organize their
memories for simple stimuli (word lists) whose temporal properties change
across different ``situations,'' analogous to how the content of real-world
experiences change across different real-world situations. We asked
participants to study and freely recall a series of word lists
(Fig.~\ref{fig:exp}). Across the different conditions in the experiment, we
varied the lists' presentation orders in different ways across lists. The
studied items (words) were designed to vary along three general dimensions:
semantic (word \textit{category}, and physical \textit{size} of the referent),
lexicographic (word \textit{length} and \textit{first letter}), and visual
(font \textit{color} and the onscreen \textit{location} of each word). In our
main manipulation conditions, we asked participants to study and recall eight
lists whose items were sorted by a target feature (e.g., word category). Next,
we asked them to study and recall an additional eight lists whose items had the
same features, but that were sorted in a random temporal order. We were
interested in how these order manipulations affected participants' recall
behaviors on early (sorted) lists, as well as how order manipulations on early
lists affected recall behaviors on later (unsorted) lists. We used a series of
control conditions as a baseline; in these control conditions all of the lists
were sorted randomly, but we manipulated the presence or absence of the visual
features. Finally, in an \textit{adaptive} experimental condition we used
participants' recall behaviors on early lists to manipulate, in real-time, the
presentation orders of subsequent lists. In this adaptive condition, we sought
to identify potential commonalities within and across participants in how
people organized their memories and how those organizational tendancies affect
overall performance.
